Follow the Flow of Rain and Paint: An Instance Optimal Contour Tree Algorithm

### 2014-06-10 paint-contour-tree.tex
more updates to painting section

\newcommand{\cut}{{\tt cut}}
\newcommand{\cont}{\psi}
\newcommand{\lift}{{\tt lift}}
\newcommand{\paint}{{\tt paint}}
\newcommand{\rain}{{\tt rain}}
\newcommand{\reeb}{R}
\newcommand{\surgery}{{\tt surgery}}
\newcommand{\touch}{T}
\newcommand{\wet}{{\tt wet}}

@@

\section{Painting to compute contour trees} \label{sec:paint}

What the previous section allows us to restrict attention to extremum dominant manifolds.
We will orient so that the extremum in question is a \emph{minimum}.
We will fix such a manifold $\MM$, with the dominant minimum $x$.

We begin by associating a color with each maximum. If the set of maxima is $S$,
this is simply an arbitrary bijection $\chi:S \mapsto [|S|]$. Imagine there being a large
can of paint of color $c(y)$ at maxima $y$. We will spill different paint from each maximum and watch it flow down.
This is analogous to the raining in the previous section, but paint is a much more viscous liquid.
\emph{So paint only flows down edges, and it does not color the interior of faces.} Furthermore, paints
do not mix, so eventually, every vertex and edge of $\MM$ gets a unique color.
%
\begin{definition} \label{def:paint} A  \emph{painting} of $\MM$ is a map $\chi:V \cup E \mapsto [|S|]$
with the following properties. (Remember that $G(\MM) = (V,E)$.)
\begin{asparaitem}
	\item The restriction $c:S \mapsto [|S|]$ is a bijection.
	\item Consider an edge $e$. There exists a descending path from some maximum $x$ to $e$
	consisting of edges in $E$, such that all edges and vertices along this path have the same color. (The maximum $x$
	is well-defined to be the unique maximum with the same color as $e$.)
\end{asparaitem}
\end{definition}
%
Note that paintings are not unique. It is straightforward to construct a painting of $\MM$ in $O(n)$ time, using a descending BFS from
each maximum. 
%
\begin{definition} \label{def:color-set} Fix a painting $\chi$, and a color $c$.
A critical point $x$ is \emph{$c$-touched} if $x$ is incident to a $c$-colored
increasing edge. The set of $c$-touched critical points is denoted $\touch(c)$.
\end{definition}
%
We will store each set $\touch(c)$ in a binomial heap. Abusing notation, $\touch(c)$ refers
both to the set and the data structure used to store it.

>>>>

\newcommand{\col}{col}
\newcommand{\mcol}{mcol}
\newcommand{\cut}{{\tt cut}}
\newcommand{\cont}{\psi}
\newcommand{\lift}{{\tt lift}}
\newcommand{\paint}{{\tt paint}}
\newcommand{\rain}{{\tt rain}}
\newcommand{\reeb}{R}
\newcommand{\rep}{rep}
\newcommand{\st}{K}
\newcommand{\surgery}{{\tt surgery}}
\newcommand{\touch}{T}
\newcommand{\update}{{\tt update}}
\newcommand{\wet}{{\tt wet}}

@@

\section{Painting to compute contour trees} \label{sec:paint}
The previous section allows us to restrict attention to extremum dominant manifolds.
We will orient so that the extremum in question is a \emph{minimum}.
We will fix such a manifold $\MM$, with the dominant minimum $u$.

We begin by associating a color with each maximum. If the set of maxima is $S$,
this is simply an arbitrary bijection $\chi:S \mapsto [|S|]$. Imagine there being a large
can of paint of color $c(y)$ at maxima $y$. We will spill different paint from each maximum and watch it flow down.
This is analogous to the raining in the previous section, but paint is a much more viscous liquid.
\emph{So paint only flows down edges, and it does not color the interior of faces.} Furthermore, paints
do not mix, so eventually, every vertex and edge of $\MM$ gets a unique color.
%
\begin{definition} \label{def:paint} A  \emph{painting} of $\MM$ is a map $\chi:V \cup E \mapsto [|S|]$
with the following properties. (Remember that $G(\MM) = (V,E)$.)
\begin{asparaitem}
	\item The restriction $c:S \mapsto [|S|]$ is a bijection.
	\item Consider an edge $e$. There exists a descending path from some maximum $x$ to $e$
	consisting of edges in $E$, such that all edges and vertices along this path have the same color. (The maximum $x$
	is well-defined to be the unique maximum with the same color as $e$.)
\end{asparaitem}
\end{definition}
%
Note that paintings are not unique. It is straightforward to construct a painting of $\MM$ in $O(n)$ time, using a descending BFS from each maximum.

\subsection{The data structures} \label{sec:struct}

{\bf The binomial heaps $\touch(c)$:} For each color $c$, we store a binomial
heap $T(c)$ of non-extremal critical points.

\begin{definition} \label{def:color-set} Fix a painting $\chi$, and a color $c$.
A critical point $x$ is \emph{touched by $c$} if $x$ is incident to a $c$-colored
increasing edge. The set of non-extremal critical points touched by $c$ is denoted $\touch(c)$.
For $x$, $\col(x)$ is the set of colors that touch $x$.
\end{definition}
%
We will store each set $\touch(c)$ in a binomial heap, keyed by the heights. 
Abusing notation, $\touch(c)$ refers
both to the set and the data structure used to store it.

\medskip
\noindent
{\bf The union-find data structure on colors:} We will repeatedly perform unions
of classes of colors, and this will be maintained as a standard union-find data structure.
For any color $c$, $\rep(c)$ denotes the representative of its class. We initialize
by simply setting $\rep(c) = c$ for all colors.

\medskip
\noindent
{\bf The stack $\st$:} This consists of non-extremal critical points with the following structure.
Each point $x \in \st$ has an associated subset of $\col(x)$, denoted $\mcol(x)$.
Both $\mcol(x)$ and its complement are stored as hash table. So lookups, inserts, and deletes
are in these sets are all constant time operations. The stack is guaranteed to satisfy 
the following invariants.
\begin{asparaitem}
	\item For every $x \in \st$: For every $c \in \mcol(x)$, $x$ is the highest element
	in $T(c)$. Furthermore, $c = \rep(c)$.
	\item Consider $x, y \in \st$ such that $y$ was pushed on $x$. There exists $c \in \col(x) \setminus
	\mcol(x)$ such that $c \in \col(y)$.
\end{asparaitem}

\subsection{Updating the stack} \label{sec:stack}

We have a simple, but critical, procedure that updates the stack $\st$ by pushing
points on to it.

\medskip
\fbox{
\begin{minipage}{0.9\textwidth}
{\bf $\update(\st)$}

\smallskip
\begin{compactenum}
	\item Let $h$ be head of $\st$.
	\item While $\col(h) \setminus \mcol(h)$ non-empty:
	\begin{compactenum}
		\item Pick some $c \in \col(h) \setminus \mcol(h)$. Replace it by $c' = \rep(c)$.
		\item Obtain highest point $x$ in $T(c')$. 
		\item If $x = h$, put $c'$ in $\mcol(h)$. Otherwise ($x \neq h$), push $x$ onto stack
		and set $\mcol(x) = \{c'\}$.
		\item Let $h$ be new head of $\st$.
	\end{compactenum}
\end{compactenum}
\end{minipage}}

\medskip
When $\update(\st)$ terminates, the head $h$ has the following property. For every $c \in \col(h)$,
$h$ is the highest point in $\touch(c)$.

\begin{claim} \label{clm:cont} Consider an increasing path from $h$ to some maxima $x$.
For any $y$ in this path, $\cont(y)$ only contains points of color $c(x)$.
Furthermore, for all vertices on $\cont(y)$, their increasing edges are colored $c(x)$.
\end{claim}

\begin{proof} Consider this increasing path, and think of moving $y$ continuously downwards from $x$
to $h$. Initially (when $y = x$), the lemma is trivially true. Consider the first $y$
when the condition fails. We have $\cont(y)$ only containing points of color $c(x)$,
but there is a vertex $z \in \cont(y)$ that has increasing edges of a different color $c'$.
All the increasing edges of $z$ that are colored $c(x)$ are contiguous (since we can construct
a contour intersecting all of them). Thus, there are two disjoint increasing neighborhoods for $z$,
and $z$ is a saddle point. But that contradicts the fact that $h$ is the highest saddle point
of color $c(x)$.

So, as $y$ decreases, for all vertices on $\cont(y)$, their increasing edges are colored $c(x)$.
This also means that $\cont(y)$ only contains points of color $c(x)$.
\end{proof}

We have important lemmas characterizing $|\col(h)|$.

\begin{lemma} \label{lem:mono} If $h$ is a merge vertex, then $|\col(h)| > 1$.
\end{lemma}

\begin{proof} Suppose not. Consider the edges incident to $h$ in circular order. There exist (at least) $2$
disjoint contiguous stretches of edges that are increasing. 
All these edges have the same color $c$, associated with some maximum $x$. Consider two neighbors $y,z$
that lie in different stretches. By the definition of a painting, there exist descending paths (along
edges) from $x$ to $y$ and $x$ to $z$. We can construct a Jordan curve $C$ from $x \rightarrow y
\rightarrow h \rightarrow z \rightarrow x$. All points in this curve except $h$ have 
heights larger than $f(h)$. By the Jordan Curve Theorem, this partitions $\MM$ into
an ``inside" and an ``outside".

One of the lower neighbors of $h$ (say $p$) is inside, and another (say $q$) is outside.
Because $h$ is a merge, neither of these can a minimum other than the dominant minimum $u$.
Hence, both $p$ and $q$ have descending paths to $u$. So there is a path from $p$
to $q$ where all heights are strictly less than $f(h)$. But this path must intersect $C$.
Contradiction.
\end{proof}

\begin{lemma} \label{lem:col} If $h$ is a merge vertex, $|\col(h)| = 2$.
If $h$ is a split vertex, $|\col(h)| = 1$.
\end{lemma}

\begin{proof} Consider one contiguous sequence of increasing edges from $h$.
There exists a contour intersecting all of these edges. By \Clm{cont}, they all
must have the same color. There are two such contiguous sequences.

If $h$ is a merge vertex, then $|\col(h)| = 2$ by \Lem{mono}. If $h$ is a split vertex,
there is a single contour passing through all increasing edges. By \Clm{cont}, they
all have the same color.
\end{proof}
